### 2021-04-13 submission/manuscript.Rmd
Edit intro based on writing group feedback

Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because of their impact on the microbiota [@Britton2014]. However, antibiotics are not the only types of medications that disrupt the microbiota [@Maier2018; @LeBastard2017; @VichVila2020]. Although, other medications have been implicated as risk factors for CDIs through epidemiological studies, whether the association is due to their impact on the microbiome is still unclear [@Oh2018; @Mora2012; @Nehra2018; @Krishna2013; @Tomkovich2019]. Many of the non-antibiotic medications associated with CDIs are known to modulate intestinal motility, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020].

Stool consistency often serves as an approximation of intestinal motility. Our group has shown that when *C. difficile* negative controls are separated into two groups based on stool consistency, there are microbiota features such as alpha diversity that overlap between samples from CDI patients and control patients with diarrhea [@Schubert2015]. These results led to our hypothesis that bacterial communities from patients experiencing diarrhea are susceptible, but may not have been exposed to *C. difficile* spores.

Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupts the intestinal microbiota in humans [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone rendered the mice susceptible to *C. difficile* infection, altered microbiota composition, reduced acetate and butyrate and altered the mucus barrier [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from *C. difficile* infections by protecting intestinal epithelial cells from the toxins produced by *C. difficile* (Ref). However, whether laxative results in more severe CDIs in mice is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT) where a healthy microbiota is administered to the patient to restore colonization. For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure. Many of the FMT studies to date rationalize the use of laxatives (Ref) based on a 1996 case study with 2 pediatric patients where the authors suggested the laxative may help flush *C. difficile* spores and toxins from the intestine [@Liacouras1996].

In the past, our group has used C57BL6 mice to characterize how antibiotics disrupt the microbiota and influence *C. difficile* susceptibility and clearance [ref]. Although, two groups have now shown PEG treatment alone renders mice susceptible to *C. difficile*, these studies have raised additional questions that should be addressed given their relevance to CDIs. Here, we used our C57BL/6 clindamycin model as a control group to characterize how long PEG-treated mice remain susceptible, whether PEG treatment results in sustained *C. difficile* colonization, and if PEG treatment post-CDI can promote *C. difficile* clearance.

>>>>

Antibiotics are a major risk factor for *Clostridioides difficile* infections (CDIs) because they disrupt microbiota colonization resistance [@Britton2014]. However, antibiotics are not the only types of medications that disrupt the microbiota [@Maier2018; @LeBastard2017; @VichVila2020]. Although, other medications (proton pump inhibitors, osmotic laxatives, antimotility agents, and opioids) have been implicated as risk or protective factors for CDIs through epidemiological studies, whether the association is due to their impact on the microbiome is still unclear [@Oh2018; @Mora2012; @Nehra2018; @Krishna2013; @Tomkovich2019].

Many of the non-antibiotic medications associated with CDIs are known to modulate gastrointestinal motility leading to either increased or decreased colonic transit time, which in turn also strongly impacts microbiota composition and function [@Vandeputte2015; @VujkovicCvijin2020]. Stool consistency often serves as an approximation of intestinal motility. Our group has shown that when *C. difficile* negative controls are separated into two groups based on stool consistency, there are shared microbiota features such as lower alpha diversity in samples from CDI patients and control patients with diarrhea compared to control samples that were *C. difficile* negative with non-diarrheal consistency [@Schubert2015]. These results led to a hypothesis that bacterial communities from patients experiencing diarrhea are susceptible to developing CDIs.

Osmotic laxatives can lead to diarrhea depending on the administered dose and temporarily disrupt the human intestinal microbiota [@Nagata2019]. The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in Miralax, Nulytely, and Golytely and is also commonly used as bowel preparation for colonoscopies. Interestingly, previous studies have shown that treating mice with PEG alone altered microbiota composition, reduced acetate and butyrate production, altered the mucus barrier, and rendered the mice susceptible to *C. difficile* infection [@Kashyap2013; @Ferreyra2014; @Tropini2018; @VanInsberghe2020]. The mucus barrier is thought to mediate protection from *C. difficile* infections by protecting intestinal epithelial cells from the toxins produced by *C. difficile* (Ref). However, whether laxative results in more severe CDIs in mice and how long mice remain colonized with *C. difficile* after challenge is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating recurrent CDIs via fecal microbiota transplant (FMT) where a healthy microbiota is administered to the patient to restore colonization. For FMTs that are delivered via colonoscopy, patients typically undergo bowel preparation by taking an osmotic laxative prior to the procedure. Many of the FMT studies to date rationalize the use of laxatives (Ref) based on a 1996 case study with 2 pediatric patients where the authors suggested in the discussion that the laxative may help flush *C. difficile* spores and toxins from the intestine [@Liacouras1996].

In the past, our group has used C57BL6 mice to characterize how antibiotics including clindamycin disrupt the microbiota and influence *C. difficile* susceptibility and clearance [ref]. Although, two groups have now shown PEG treatment alone renders mice susceptible to *C. difficile*, these studies have raised additional questions regarding the dynamics and severity of infection as well as the role of laxative treatment in *C. difficile* clearance that should be addressed to better inform how we think about laxatives in the context of CDIs. Here, we used our C57BL/6 clindamycin model as a control group to characterize how long PEG-treated mice remain susceptible, whether PEG treatment results in more severe CDI and sustained *C. difficile* colonization, and if PEG treatment post-CDI can promote *C. difficile* clearance.

### 2021-04-13 submission/manuscript.tex
Edit intro based on writing group feedback

Antibiotics are a major risk factor for \emph{Clostridioides difficile}
infections (CDIs) because of their impact on the microbiota (1).
However, antibiotics are not the only types of medications that disrupt
the microbiota (2--4). Although, other medications have been implicated
as risk factors for CDIs through epidemiological studies, whether the
association is due to their impact on the microbiome is still unclear
(5--9). Many of the non-antibiotic medications associated with CDIs are
known to modulate intestinal motility, which in turn also strongly
impacts microbiota composition and function (10, 11).

Stool consistency often serves as an approximation of intestinal
motility. Our group has shown that when \emph{C. difficile} negative
controls are separated into two groups based on stool consistency, there
are microbiota features such as alpha diversity that overlap between
samples from CDI patients and control patients with diarrhea (12). These
results led to our hypothesis that bacterial communities from patients
experiencing diarrhea are susceptible, but may not have been exposed to
\emph{C. difficile} spores.

Osmotic laxatives can lead to diarrhea depending on the administered
dose and temporarily disrupts the intestinal microbiota in humans (13).
The ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found
in Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone rendered the mice susceptible to
\emph{C. difficile} infection, altered microbiota composition, reduced
acetate and butyrate and altered the mucus barrier (14--17). The mucus
barrier is thought to mediate protection from \emph{C. difficile}
infections by protecting intestinal epithelial cells from the toxins
produced by \emph{C. difficile} (Ref). However, whether laxative results
in more severe CDIs in mice is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating
recurrent CDIs via fecal microbiota transplant (FMT) where a healthy
microbiota is administered to the patient to restore colonization. For
FMTs that are delivered via colonoscopy, patients typically undergo
bowel preparation by taking an osmotic laxative prior to the procedure.
Many of the FMT studies to date rationalize the use of laxatives (Ref)
based on a 1996 case study with 2 pediatric patients where the authors
suggested the laxative may help flush \emph{C. difficile} spores and
toxins from the intestine (18).

In the past, our group has used C57BL6 mice to characterize how
antibiotics disrupt the microbiota and influence \emph{C. difficile}
susceptibility and clearance {[}ref{]}. Although, two groups have now
shown PEG treatment alone renders mice susceptible to \emph{C.
difficile}, these studies have raised additional questions that should
be addressed given their relevance to CDIs. Here, we used our C57BL/6
clindamycin model as a control group to characterize how long
PEG-treated mice remain susceptible, whether PEG treatment results in
sustained \emph{C. difficile} colonization, and if PEG treatment
post-CDI can promote \emph{C. difficile} clearance.

>>>>

Antibiotics are a major risk factor for \emph{Clostridioides difficile}
infections (CDIs) because they disrupt microbiota colonization
resistance (1). However, antibiotics are not the only types of
medications that disrupt the microbiota (2--4). Although, other
medications (proton pump inhibitors, osmotic laxatives, antimotility
agents, and opioids) have been implicated as risk or protective factors
for CDIs through epidemiological studies, whether the association is due
to their impact on the microbiome is still unclear (5--9).

Many of the non-antibiotic medications associated with CDIs are known to
modulate gastrointestinal motility leading to either increased or
decreased colonic transit time, which in turn also strongly impacts
microbiota composition and function (10, 11). Stool consistency often
serves as an approximation of intestinal motility. Our group has shown
that when \emph{C. difficile} negative controls are separated into two
groups based on stool consistency, there are shared microbiota features
such as lower alpha diversity in samples from CDI patients and control
patients with diarrhea compared to control samples that were \emph{C.
difficile} negative with non-diarrheal consistency (12). These results
led to a hypothesis that bacterial communities from patients
experiencing diarrhea are susceptible to developing CDIs.

Osmotic laxatives can lead to diarrhea depending on the administered
dose and temporarily disrupt the human intestinal microbiota (13). The
ubiquitous osmotic laxative, polyethylene glycol (PEG) 3350 is found in
Miralax, Nulytely, and Golytely and is also commonly used as bowel
preparation for colonoscopies. Interestingly, previous studies have
shown that treating mice with PEG alone altered microbiota composition,
reduced acetate and butyrate production, altered the mucus barrier, and
rendered the mice susceptible to \emph{C. difficile} infection (14--17).
The mucus barrier is thought to mediate protection from \emph{C.
difficile} infections by protecting intestinal epithelial cells from the
toxins produced by \emph{C. difficile} (Ref). However, whether laxative
results in more severe CDIs in mice and how long mice remain colonized
with \emph{C. difficile} after challenge is unclear.

Beyond susceptibility, PEG is also relevant in the context of treating
recurrent CDIs via fecal microbiota transplant (FMT) where a healthy
microbiota is administered to the patient to restore colonization. For
FMTs that are delivered via colonoscopy, patients typically undergo
bowel preparation by taking an osmotic laxative prior to the procedure.
Many of the FMT studies to date rationalize the use of laxatives (Ref)
based on a 1996 case study with 2 pediatric patients where the authors
suggested in the discussion that the laxative may help flush \emph{C.
difficile} spores and toxins from the intestine (18).

In the past, our group has used C57BL6 mice to characterize how
antibiotics including clindamycin disrupt the microbiota and influence
\emph{C. difficile} susceptibility and clearance {[}ref{]}. Although,
two groups have now shown PEG treatment alone renders mice susceptible
to \emph{C. difficile}, these studies have raised additional questions
regarding the dynamics and severity of infection as well as the role of
laxative treatment in \emph{C. difficile} clearance that should be
addressed to better inform how we think about laxatives in the context
of CDIs. Here, we used our C57BL/6 clindamycin model as a control group
to characterize how long PEG-treated mice remain susceptible, whether
PEG treatment results in more severe CDI and sustained \emph{C.
difficile} colonization, and if PEG treatment post-CDI can promote
\emph{C. difficile} clearance.
